---
title: "Pan_homo_carotid_scPrep"
output: html_document
date: "2024-04-01"
editor_options: 
  chunk_output_type: console
---
# Load libraries
```{r}
setwd("Human/")

library(Seurat)
library(tidyverse)
library(data.table)
library(scDblFinder)
library(celda)
library(sctransform)
library(cluster)
library(readr)
source("../scRNA_processing_utils.R") # with the modifications from https://github.com/MillerLab-CPHG/Human_athero_scRNA_meta/blob/main/utils/scRNA_processing_utils.R

# Set seed for doublet removal reproducibility
set.seed(1)

```

# Sample 1: GSM4705589 RPE004 (Caucasian, Male, 83, Symptomatic)

## Prepare the count matrix

```{r}
# Load Matrix (or use ReadMtx {Seurat})
rpe004_matrix = data.table::fread("Pan_carotids_scRNA/GSM4705589/GSM4705589_RPE004_matrix.txt", sep="\t", header = TRUE)

rownames(rpe004_matrix) = rpe004_matrix$gene
head(rownames(rpe004_matrix))
dim(rpe004_matrix)

# Make sparse matrix and remove "gene" column
rpe004_sparse_mtx = as.sparse(rpe004_matrix)
rpe004_sparse_mtx = rpe004_sparse_mtx[, -1]

# 15796 genes x 2614 cells
dim(rpe004_sparse_mtx)

```

## Create Seurat obj
```{r}
#Pre-processing (Create Seurat obj containing doublets)
# Dims of seurat object: 15179 genes x 2614 cells
rpe004_seurat_sct = CreateSeuratObject(counts = rpe004_sparse_mtx, 
                                      project = "pan_rpe004_carotid", 
                                      min.cells = 10, 
                                      min.features = 200)

#  before removing doublets
# SCT normalize
rpe004_seurat_sct = Seurat_SCT_process(rpe004_seurat_sct, seurat_filter = TRUE,
                                      sample_id = "pan_rpe004", 
                                      study_name = "pan_et_al")

###########
### Finding Best Resolution for integrated object

resolutions_sil_df = calc_sil_scores_for_res(rpe004_seurat_sct, clustering_res = c(0.5, 0.6, 0.7, 0.9, 0.75 ,0.8, 1.0,1.1, 1.4))

max_avg_sil_score(resolutions_sil_df) # 0.6 is Max
##########

# Finding the right resolution is quite important for doublet detection and removal
rpe004_seurat_sct =  FindClusters(rpe004_seurat_sct, resolution = 0.5) #As suggested by authors

# Visualize clusters
p1_before_QC = DimPlot(rpe004_seurat_sct, reduction = "umap", label = TRUE) +
  ggtitle("Before removing doublets") + NoAxes()
ggsave(filename = "Pan_carotids_scRNA/Pan_rpe004_before_QC.png", width = 7, height = 8)

```

## Detect doublets & ambient RNA

```{r}
# Detect doublets with scDblFinder
# Called 3 doublets
rpe004_doublet_ids = scDblFinder_clusters(rpe004_seurat_sct, nrep = 3)
length(rpe004_doublet_ids)

# First filtering step: Use the vector of consensus doublet IDs output below
rpe004_singlets = setdiff(colnames(rpe004_sparse_mtx), rpe004_doublet_ids)
length(rpe004_singlets)

# New dims after removing doublets: 15174 genes x 2606 cells
rpe004_seurat_sct = CreateSeuratObject(counts = rpe004_sparse_mtx[, rpe004_singlets], 
                                      project = "pan_rpe004", 
                                      min.cells = 10, 
                                      min.features = 200)

# Clean counts matrix (without doublets) from ambient RNA using the decontX wrapper
rpe004_seurat_sct =  decontX_remove(rpe004_seurat_sct)
head(rpe004_seurat_sct@assays$RNA@counts)

```

```{r}
# after removing doublets
# SCT normalize
rpe004_seurat_sct = Seurat_SCT_process(rpe004_seurat_sct, seurat_filter = TRUE,
                                      sample_id = "pan_rpe004", 
                                      study_name = "pan_et_al")

# Finding the right resolution is quite important for doublet detection and removal
rpe004_seurat_sct =  FindClusters(rpe004_seurat_sct, resolution = 0.5)

p1_after_QC = DimPlot(rpe004_seurat_sct, reduction = "umap", label = TRUE) +
  ggtitle("After removing doublets + decontX") + NoAxes()
ggsave(filename = "Pan_carotids_scRNA/Pan_rpe004_after_QC.png", width = 7, height = 8)

FeaturePlot(rpe004_seurat_sct, features = c("MYH11", "CNN1", "TNFRSF11B", "ACTA2", 
                                           "CRTAC1", "LUM"), order = TRUE,  pt.size = 0.1)

FeaturePlot(rpe004_seurat_sct, features = c("CD68", "CD14", "CD163", "APOE", 
                                           "S100A8", "IL1B"), order = TRUE,  pt.size = 0.1)

FeaturePlot(rpe004_seurat_sct, features = c("MYH11", "CNN1", "FHL5", "FOXC1"), order = TRUE,  pt.size = 0.1)

saveRDS(rpe004_seurat_sct, file = "Pan_carotids_scRNA/GSM4705589_pan_rpe004_processed_seurat_obj.rds")
```

# Sample 2: GSM4705590 RPE005 (Caucasian, Male, 67, Asymptomatic)
## Prepare the count matrix

```{r}
# Load Matrix
rpe005_matrix = fread("Pan_carotids_scRNA/GSM4705590/GSM4705590_RPE005_matrix.txt", sep="\t", header = TRUE)

rownames(rpe005_matrix) = rpe005_matrix$gene
head(rownames(rpe005_matrix))
dim(rpe005_matrix)

# Make sparse matrix and remove "gene" column
rpe005_sparse_mtx = as.sparse(rpe005_matrix)
rpe005_sparse_mtx = rpe005_sparse_mtx[, -1]

# 17397 genes x 3486 cells
dim(rpe005_sparse_mtx)
```

## Create Seurat obj
```{r}
#Pre-processing (Create Seurat obj containing doublets)
# Dims of seurat object: 16670 genes x 3486 cells
rpe005_seurat_sct = CreateSeuratObject(counts = rpe005_sparse_mtx, 
                                       project = "pan_rpe005", 
                                       min.cells = 10, 
                                       min.features = 200)

#  before removing doublets
# SCT normalize
rpe005_seurat_sct = Seurat_SCT_process(rpe005_seurat_sct, seurat_filter = TRUE,
                                       sample_id = "pan_rpe005", 
                                       study_name = "pan_et_al")

###########
### Finding Best Resolution for integrated object

resolutions_sil_df = calc_sil_scores_for_res(rpe005_seurat_sct, clustering_res = c(0.5, 0.6, 0.7, 0.9, 0.75 ,0.8, 1.0,1.1, 1.4))

max_avg_sil_score(resolutions_sil_df) # 0.6 is Max
##########

# Finding the right resolution is quite important for doublet detection and removal
rpe005_seurat_sct =  FindClusters(rpe005_seurat_sct, resolution = 0.5)   

# Visualize clusters
p2_before_QC = DimPlot(rpe005_seurat_sct, reduction = "umap", label = TRUE) +
  ggtitle("Before removing doublets") + NoAxes()
ggsave(filename = "Pan_carotids_scRNA/Pan_rpe005_before_QC.png", width = 7, height = 8)

```

## Detect doublets & ambient RNA

```{r}
# Detect doublets with scDblFinder
# Called 8 doublets
rpe005_doublet_ids = scDblFinder_clusters(rpe005_seurat_sct, nrep = 3)
length(rpe005_doublet_ids)

# First filtering step: Use the vector of consensus doublet IDs output below
rpe005_singlets = setdiff(colnames(rpe005_sparse_mtx), rpe005_doublet_ids)
length(rpe005_singlets)

# New dims after removing doublets: 16640 genes x 3436 cells
rpe005_seurat_sct = CreateSeuratObject(counts = rpe005_sparse_mtx[, rpe005_singlets], 
                                       project = "pan_rpe005", 
                                       min.cells = 10, 
                                       min.features = 200)

# Clean counts matrix (without doublets) from ambient RNA using the decontX wrapper
rpe005_seurat_sct =  decontX_remove(rpe005_seurat_sct)
head(rpe005_seurat_sct@assays$RNA@counts)
```

```{r}
# after removing doublets
# SCT normalize
rpe005_seurat_sct = Seurat_SCT_process(rpe005_seurat_sct, seurat_filter = TRUE,
                                       sample_id = "pan_rpe005", 
                                       study_name = "pan_et_al")

# Finding the right resolution is quite important for doublet detection and removal
rpe005_seurat_sct =  FindClusters(rpe005_seurat_sct, resolution = 0.5) #Max score is 0.5   

p2_after_QC = DimPlot(rpe005_seurat_sct, reduction = "umap", label = TRUE) +
  ggtitle("After removing doublets + decontX") + NoAxes()
ggsave(filename = "Pan_carotids_scRNA/Pan_rpe005_after_QC.png", width = 7, height = 8)

FeaturePlot(rpe005_seurat_sct, features = c("MYH11", "CNN1", "TNFRSF11B", "ACTA2", 
                                            "CRTAC1", "LUM"), order = TRUE,  pt.size = 0.1)

FeaturePlot(rpe005_seurat_sct, features = c("CD68", "CD14", "CD163", "APOE", 
                                            "S100A8", "IL1B"), order = TRUE,  pt.size = 0.1)

saveRDS(rpe005_seurat_sct, file = "Pan_carotids_scRNA/GSM4705590_pan_rpe005_processed_seurat_obj.rds")

```

# Sample 3: GSM4705591 RPE006 (Caucasian, Female, 76, Asymptomatic)

## Prepare the count matrix

```{r}
# Load Matrix
rpe006_matrix = fread("Pan_carotids_scRNA/GSM4705591/GSM4705591_RPE006_matrix.txt", sep="\t", header = TRUE)

rownames(rpe006_matrix) = rpe006_matrix$gene
head(rownames(rpe006_matrix))
dim(rpe006_matrix)

# Make sparse matrix and remove "gene" column
rpe006_sparse_mtx = as.sparse(rpe006_matrix)
rpe006_sparse_mtx = rpe006_sparse_mtx[, -1]

# 15687 genes x 2767 cells
dim(rpe006_sparse_mtx)
```

## Create Seurat obj

```{r}
# Pre-processing (Create Seurat obj containing doublets)
# Dims of seurat object: 15445 genes x 2767 cells
rpe006_seurat_sct = CreateSeuratObject(counts = rpe006_sparse_mtx, 
                                       project = "pan_rpe006", 
                                       min.cells = 10, 
                                       min.features = 200)
#  before removing doublets
# SCT normalize
rpe006_seurat_sct = Seurat_SCT_process(rpe006_seurat_sct, seurat_filter = TRUE,
                                       sample_id = "pan_rpe006", 
                                       study_name = "pan_et_al")

###########
### Finding Best Resolution for integrated object

resolutions_sil_df = calc_sil_scores_for_res(rpe006_seurat_sct, clustering_res = c(0.5, 0.6, 0.7, 0.9, 0.75 ,0.8, 1.0,1.1, 1.4))

max_avg_sil_score(resolutions_sil_df) # 0.6 is Max
##########

# Finding the right resolution is quite important for doublet detection and removal
rpe006_seurat_sct =  FindClusters(rpe006_seurat_sct, resolution = 0.5)   

# Visualize clusters
p3_before_QC = DimPlot(rpe006_seurat_sct, reduction = "umap", label = TRUE) +
  ggtitle("Before removing doublets") + NoAxes()

ggsave(filename = "Pan_carotids_scRNA/Pan_rpe006_before_QC.png", width = 7, height = 8)

```

## Detect doublets & ambient RNA

```{r}
# Detect doublets with scDblFinder
rpe006_doublet_ids = scDblFinder_clusters(rpe006_seurat_sct, nrep = 3)
length(rpe006_doublet_ids)

# First filtering step: Use the vector of consensus doublet IDs output below
rpe006_singlets = setdiff(colnames(rpe006_sparse_mtx), rpe006_doublet_ids)
length(rpe006_singlets)

# New dims after removing doublets: 15406 genes x 2730 cells
rpe006_seurat_sct = CreateSeuratObject(counts = rpe006_sparse_mtx[, rpe006_singlets], 
                                       project = "pan_rpe006", 
                                       min.cells = 10, 
                                       min.features = 200)

# Clean counts matrix (without doublets) from ambient RNA using the decontX wrapper
rpe006_seurat_sct =  decontX_remove(rpe006_seurat_sct)
head(rpe006_seurat_sct@assays$RNA@counts)
```

```{r}
# after removing doublets
# SCT normalize

rpe006_seurat_sct = Seurat_SCT_process(rpe006_seurat_sct, seurat_filter = TRUE,
                                       sample_id = "pan_rpe006", 
                                       study_name = "pan_et_al")

# Finding the right resolution is quite important for doublet detection and removal
rpe006_seurat_sct =  FindClusters(rpe006_seurat_sct, resolution = 0.5) #Max is 0.5 

p3_after_QC = DimPlot(rpe006_seurat_sct, reduction = "umap", label = TRUE) +
  ggtitle("After removing doublets + decontX") + NoAxes()
ggsave(filename = "Pan_carotids_scRNA/Pan_rpe006_after_QC.png", width = 7, height = 8)

FeaturePlot(rpe006_seurat_sct, features = c("MYH11", "CNN1", "TNFRSF11B", "ACTA2", 
                                            "CRTAC1", "VCAM1"), order = TRUE,  pt.size = 0.1)

FeaturePlot(rpe006_seurat_sct, features = c("CD68", "CD14", "CD163", "APOE", 
                                            "S100A8", "IL1B"), order = TRUE,  pt.size = 0.1)

saveRDS(rpe006_seurat_sct, file = "Pan_carotids_scRNA/GSM4705591_pan_rpe006_processed_seurat_obj.rds")

```

